### 2015-01-20 manuscript/SpatPred3.tex
SpatPred draft to Peter and Devin for first time - doesn't have Steller example integrated into text yet

\begin{center}
Avoiding extrapolation bias when using statistical models to make ecological prediction
\bigskip\\
\normalsize
{\sc Paul B. Conn$^{1,}$\footnotemark[2] and
Devin S. Johnson$^1$, Peter L. Boveng$^1$ }\smallskip\\
$^1${\em National Marine Mammal Laboratory, NOAA, National Marine Fisheries Service,
Alaska Fisheries Science Center, 7600 Sand Point Way NE, Seattle,
WA 98115 USA }\\ \medskip
\end{center}
\footnotetext[2]{Email: paul.conn@noaa.gov}

@@

In ecology and conservation, a common goal is to make predictions about an unsampled random variable given a limited sample from the target population.  For instance, given a model ($\mathcal{M}$), estimated parameters ($\hat{\boldsymbol{\theta}}$), and a covariate vector ${\bf x}_i$, we often desire to predict a new observation $y_{new}$ at $i$.  For instance, we might use a generalized linear model \citep{McCullaghNelder1989} or one of its extensions to predict species density or occurrence in a new location, or to predict the future trend of a population.

Early in their training, ecologists and statisticians are warned against extrapolating statistical relationships
past the range of observed data.  This caution is easily interpreted in the context of single-variable linear regression analysis; one should be cautious in using the fitted relationship to make predictions at some new point $y_{new}$ whenever $x_{new} < \min({\bf x})$ or $x_{new} > \max({\bf x})$.  But what about more complicated situations
where there are multiple explanatory variables, or when one uses a spatial regression model to account for the residual spatial autocorrelation that is inevitably present in patchy ecological data \citep{LichsteinEtAl2002}?  How reliable are spatially- or temporally-explicit predictions in sophisticated models for animal abundance and occurrence?

@@

Our aim in this paper is to investigate extrapolation bias in the generalized linear model and its extensions, including generalized additive models \citep[GAMs;][]{HastieTibshirani1999,Wood2006} and spatial, temporal, or spatio-temporal regression models (STRMs).  In particular, we exploit some of the same ideas used in multiple linear regression regarding leverage and outliers \citep{Cook1979} to operationally define ``extrapolation" as making predictions that occur outside of a generalized independent variable hull (gIVH) of observed data points. Application of the gIVH and related criterion (e.g. prediction variance) can provide intuition regarding the reliability of predictions in unobserved locations, and can aid in model construction.  Also, since the gIVH can be constructed solely with knowledge of sampled locations and explanatory covariates (i.e., it does not necessarily require any observed response variables), it can also be used to help guide survey design. We illustrate use of the gIVH on a simulated occupancy dataset, on a species distribution model (SDM) for ribbon seals in the eastern Bering Sea, and on a population trend model for Steller Sea Lions ({\it Phoca fasciata}).

@@

We simulated two sets of design data to help illustrate application of the IVH (Fig. \ref{fig:IVH}).  In simple linear regression with one predictor variable, predictions on a hypothetical response variable obtained at covariate values below the lowest observed value or above the highest observed value are primarily outside of the IVH.  We suspect this result conforms to most ecologists intuition about what constitutes ``extrapolating past the observed data."  However, fitting a quadratic model exhibits more nuance; if there is a large gap between design points, it is entirely possible that intermediate covariate values will also be outside of the IVH and thus more likely to result in problematic predictions.  Fitting a model with two covariates and both linear and quadratic effects, the shape of the IVH is somewhat more irregular, and even includes a hole in the middle of the surface when interactions are modeled (Fig. \ref{fig:IVH}).  These simple examples highlight the sometimes counterintuitive nature of the predictive inference problem, a problem that can only become worse as models with more dimensions are contemplated (including those with temporal or spatial structure).  Fortunately, the ideas behind the IVH provide a potential way forward.

Cook's (1979) formulation for the IVH is particular to linear regression analysis, which assumes iid Gaussian error. Thus, it is not directly applicable to generalized models, such as those including alternative error distributions (e.g., Poisson, binomial) or spatial random effects.  Further, the hat matrix is not necessarily well defined for models with more general spatial structure. However, since the hat matrix is proportional to prediction variance, Cook (1979) notes that design points with maximum prediction variance will be located on the boundary of the IVH.  We therefore define a generalized independent variable hull (gIVH) as the set of all points ${\bf x}$ (note that ${\bf x}$ can include both observed and unobserved design points) such that
\begin{linenomath*}
\begin{equation}
  \label{eq:gIVH}
  {\rm var}({\bf y}_x | {\bf x}) \le {\rm max} [{\rm var}({\bf y}_X | {\bf X})],
\end{equation}
\end{linenomath*}
where ${\bf Y}_x$ correspond to predictions at ${\bf x}$, and ${\bf X}$ and ${\bf Y}_X$ denote observed design points and predictions at $X$, respectively.

Generalizations of the linear model are often written in the form
\begin{linenomath*}
\begin{equation}
  Y_i \sim f_Y(g(\mu_i)),
\end{equation}
\end{linenomath*}
where $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives an inverse link function, and
$\mu_i$ is a predictor.  For many such generalizations, it is possible to specify the $\mu_i$ as
\begin{linenomath*}
\begin{equation}
  \boldsymbol{\mu} = {\bf X}_{aug}\boldsymbol{\beta}_{aug} + \boldsymbol{\epsilon},
  \label{eq:mu-aug}
\end{equation}
\end{linenomath*}
where the $\boldsymbol{\epsilon}$ represent Gaussian error, ${\bf X}_{aug}$ denotes an augmented design matrix, and $\boldsymbol{\beta}_{aug}$ denote an augmented vector of parameters.  For instance, in a spatial model, $\boldsymbol{\beta}_{aug}$ might include both fixed effect parameters and spatial random effects in a reduced dimension subspace (see Appendix A for examples of how numerous types of models can be written in this form).

When models are specified as in Eq. \ref{eq:mu-aug}, we can write prediction variance generically as
\begin{linenomath*}
\begin{equation}
  \label{eq:var-mu}
  {\rm var}(\boldsymbol{\mu}_x | {\bf x})= {\bf x} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}) {\bf x}^\prime,
\end{equation}
\end{linenomath*}
where it is understood that the exact form of ${\bf x}$ and ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the model chosen (i.e., GLM, GAM, or STRM; Appendix A). This expression for prediction variance is on the linear predictor scale; if a non-identity link function is used, we can use the delta method \citep{Dorfman1938,VerHoef2012} to approximate prediction variance on the real scale (i.e. the scale in which data are measured) as
\begin{linenomath*}
\begin{eqnarray*}
  \label{eq:var-mu}
  {\rm var}(g(\mu_x) | {\bf x}) & = & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x | {\bf x})\boldsymbol{\Delta}^\prime, 
%  \boldsymbol{\Delta} & = & \left[ \begin{array}{cccc}
%      \frac{\partial g(\mu_1)}{\partial \mu_1} & \frac{\partial g(\mu_1)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_1)}{\partial \mu_n} \\
%      \frac{\partial g(\mu_2)}{\partial \mu_1} & \frac{\partial g(\mu_2)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_2)}{\partial \mu_n} \\
%      \vdots & \vdots & \ddots & \vdots \\
%      \frac{\partial g(\mu_n)}{\partial \mu_1} & \frac{\partial g(\mu_n)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_n)}{\partial \mu_n} \\
%  \end{array} \right].
\end{eqnarray*}
\end{linenomath*}
where $\boldsymbol{\Delta}$ is a matrix of partial derivatives $\partial g(\mu_r)/\partial \mu_c$ ($r$ and $c$ denoting rows and columns of $\boldsymbol{\Delta}$, respectively).  Under common univariate link functions (e.g. log, logit, probit), $\boldsymbol{\Delta}$ has a diagonal form, while for multivariate links (e.g. multinomial logit) $\boldsymbol{\Delta}$ will be dense.

%For GLMs, prediction variance is proportional to the original hat matrix ${\bf V}_{LR}$ (at %least on the linear predictor scale), and Eq. \ref{eq:IVH} may be used directly.  For other %models, we have
%\begin{linenomath*}
%\begin{equation}
%  \label{eq:var_Beta}
%  {\rm var}(\hat{\boldsymbol{\beta}}_{aug})=\left[ \begin{array}{cc}
%      \boldsymbol{\Sigma}_\beta & {\bf 0}\\
%      {\bf 0} & \boldsymbol{\Sigma}_\alpha
%  \end{array} \right],
%\end{equation}
%\end{linenomath*}
%where we simply replace $\boldsymbol{\beta}_{aug}$ with the augmented vector that includes %both regression and spatial or smooth parameters (i.e. $\boldsymbol{\beta}$ and %$\boldsymbol{\alpha}$; Appendix S1).

The exact form of ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ differs depending on the underlying model structure and estimation procedure.  In the following treatment, we shall focus on Bayesian analysis; although it is not necessarily needed to fit GLMs and GAMs, it helps make STRMs more computationally tractable and puts all of the different models into a common analysis framework.  This approach requires specifying priors for model parameters, and prior parameters may appear in expressions for prediction variance.  Judicious choices of priors can at times limit the influence of priors on calculation of the gIVH.  For example, if we specify the prior on the vector of regression coefficients to be $[\boldsymbol{\beta}] = \textrm{MVN}({\bf 0},(\tau_\beta X^\prime X)^{-1})$ where $\tau_\beta$ is a fixed constant, a Bayesian implementation of a GLM still results in the gIVH specified in Eq. \ref{eq:IVH}.  The ability to use Eq. 5 directly is quite useful, as we only need to know the explanatory variables to be able to diagnose whether predictions lie in or out of the gIVH. However, the other models considered here (GAMs and STRMs) typically require estimation of one or more precision or autocorrelation parameters.   In these cases, one solution is to impose prior distributions and integrate over these additional parameters (call them $\boldsymbol{\theta}$) to obtain an expectation for ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$.  Using the law of the unconscious statistician, we have
\begin{linenomath*}
\begin{equation}
  \label{eq:E_var_Beta}
  {\rm E}({\rm var}(\hat{\boldsymbol{\beta}}_{aug}))  = \int_{\boldsymbol{\theta}} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}|\boldsymbol{\theta}) [\boldsymbol{\theta}] d \boldsymbol{\theta}.
\end{equation}
\end{linenomath*}
Here, and throughout the paper, we use the bracket notation to indicate a probability density function; e.g., $[\boldsymbol{\theta}] $ denotes the joint prior distribution for $\boldsymbol{\theta}$. If observations have already been gathered, one could also consider integrating over the marginal posterior distribution using MCMC:
\begin{linenomath*}
\begin{equation}
  \label{eq:E_var_Beta_post}
  {\rm E}({\rm var}(\hat{\boldsymbol{\beta}}_{aug}) | {\bf Y})  = \int_{\boldsymbol{\theta}} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}|\boldsymbol{\theta}) [\boldsymbol{\theta}| {\bf Y}] d \boldsymbol{\theta}.
\end{equation}
\end{linenomath*}
Some care must be taken with prior distributions if Eq. \ref{eq:E_var_Beta} is used for calculation of prediction variance.  For instance, standard ``non-informative" or ``flat" priors may place substantial mass on implausible values.  When using the gIVH or prediction variance to help guide sampling design \citep[see][for an example using prediction variance]{DiggleLophaven2006}, we suggest using informative prior distributions.

We propose to use the gIVH in much the same manner as Cook (1979).  In particular, we use the gIVH to differentiate whether spatial predictions are interpolations (predictive design points lying inside the gIVH) or extrapolations (predictive design points lying outside the gIVH).  The gIVH, together with prior-integrated prediction variance, seem ideally situated to diagnosing potential extrapolation issues as it does not necessarily need to rely on gathered response data. Thus, one can examine whether or not prediction points lie within the gIVH without ever collecting response data there.  Further, one can compare prediction variance in places one has data to places where predictions are desired to gauge the relative amount of information that predictions are being based on.

@@

@@

We conducted a simulation study to investigate whether the gIVH (and accompanying prediction variance) was useful in diagnosing prediction biases when analyzing animal count data. For each of 100 simulations, we generated animal abundance over a $30 \times 30$ grid assuming that animal density was homogeneous in each grid cell.  Animal abundance was generated as a function of four hypothetical habitat covariates (Appendix B).   For each simulated landscape, we randomly selected 60 grid cells for sampling, assuming that sampling quadrats encompassed 10\% of the area of each selected grid cell.  For ease of presentation and analysis, we assumed detection probability was 100\% in each quadrat.  Once animal counts were simulated, three different estimation models were fitted to the data: a GLM, a GAM, and an STRM (Appendix B).  The GLM and STRM expressed log-density as a function of linear and quadratic covariate effects, while the GAM used a kernel smoother with 4 knots for each covariate (Appendix B).  Each model was provided with three of the four covariates used to generate the data.

For each simulated data set and model structure, we calculated (1) $\text{gIVH}_{int}$, the gIVH resulting from integrating over prior parameter variance (i.e. using Eq. \ref{eq:E_var_Beta} with informative priors), and (2) $\text{gIVH}_{post}$, the gIVH arising from posterior predictive variance (i.e. posterior variance when data are actually analyzed).  We then calculated posterior predictions of animal abundance within and outside of each gIVH in order to gauge bias as a function of whether or not inference is constrained to the gIVH.  Specifically, the performance of the $\text{gIVH}_{int}$ criterion in diagnosing extrapolation bias is a test of whether it is useful in designing animal count surveys, while the performance of $\text{gIVH}_{post}$ may help decide its utility in limiting the scope of landscape-based inference once data have been collected and analyzed.

\subsection{Ribbon seal SDM}
As part of an international effort, researchers with the U.S. National Marine Fisheries Service conducted aerial surveys over the eastern Bering Sea in 2012 and 2013.  Agency scientists used infrared video to detect seals that were on ice, and simultaneous automated digital photographs provided information on species identity. Here, we use spatially referenced count data from photographed ribbon seals, {\it Phoca fasciata} (Fig. \ref{fig:ribbon}) on a subset of 10 flights flown over the Bering Sea in April 2012.  These flights were limited to a one week period so that both sea ice conditions and seal distributions could be assumed to be static.

Our objective with this dataset will be to model seal counts on transects through 25km by 25km grid cells as a function of habitat covariates and possible spatial autocorrelation. Estimates of apparent abundance can then be obtained by summing predictions across grid cells. Figure \ref{fig:flights} shows the transects flown and the number of ribbon seals encountered in each cell, and Figure \ref{fig:covs} show explanatory covariates gathered to help predict ribbon seal abundance.  These data are described in fuller detail by \citet{ConnEtAl2014}, who extend the modeling framework of STRMs to account for incomplete detection and species misidentification errors \citep[see e.g.][]{ConnEtAl2014}.  Since our focus in this paper is on illustrating spatial modeling concepts, we devote our efforts to the comparably easier problem of estimating apparent abundance (i.e., uncorrected for vagaries of the detection process).

Inspection of ribbon seal data (Fig. \ref{fig:flights}) immediately reveals a potential
issue with spatial prediction: abundance of ribbon seals appears to be maximized in the southern and/or southeast quadrant of the surveyed area.  Predicting abundance in areas further south and east may thus prove problematic.  To illustrate, let $Y_i$ denote the ribbon seal count ($Y_i$) obtained in sampled grid cell $i$.  Suppose that counts arise according to a log-Gaussian Cox process, such that
\begin{eqnarray}
  \label{eqn:Cox}
  Y_i \sim {\rm Poisson}(\lambda_i) \mbox{ and } \\
  \log(\lambda_i) = \log(P_i) + {\bf X}_i \boldsymbol{\beta} + \eta_i + \epsilon_i, \nonumber
\end{eqnarray}
where $P_i$ gives the proportion of area photographed in grid cell $i$ (recall also that ${\bf X}_i$ denotes a vector covariates for cell $i$, $\boldsymbol{\beta}$ are regression coefficients, $\eta_i$ represents a spatially autocorrelated random effect, and $\epsilon_i$ is normally distributed $iid$ error).

We could fit any number of predictive models to these data, but we start with a simple
generalized linear model where we ignore the spatial random effect, $\eta_i$, and use the full suite of predictor covariates (Fig. \ref{fig:covs}) to fit Eq. \ref{eqn:Cox} to our data. In particular, we fit a model with linear effects of all predictor variables, and with an additional quadratic term for ice concentration \citep[seal density is often maximized at an intermediate value of ice concentration; see][]{VerHoef2013,ConnEtAl2014}. To enable comparison with more complicated types of models, we formulated a generalized Bayesian strategy for parameter estimation (see Appendix S1).  For simplicity, we generated posterior predictions of ribbon seal abundance across the landscape as
\begin{equation}
  N_i \sim {\rm Poisson}(A_i \lambda_i),
\end{equation}
where $A_i$ gives the proportion of suitable habitat in cell $i$ (ribbon seals do not inhabit land masses).

Fitting this model to our data,

>>>>

\begin{center}
On extrapolating past the range of observed data when making statistical predictions in ecology
\bigskip\\
\normalsize
{\sc Paul B. Conn$^{1,}$\footnotemark[2] and
Devin S. Johnson$^1$, Peter L. Boveng$^1$ }\smallskip\\
$^1${\em National Marine Mammal Laboratory, NOAA, National Marine Fisheries Service,
Alaska Fisheries Science Center, 7600 Sand Point Way NE, Seattle,
WA 98115 USA }\\ \medskip
\end{center}
\footnotetext[2]{Email: paul.conn@noaa.gov}

@@

In ecology and conservation, a common goal is to make predictions about an unsampled random variable given a limited sample from the target population.  For instance, given a model ($\mathcal{M}$), estimated parameters ($\hat{\boldsymbol{\theta}}$), and a covariate vector ${\bf x}_s$, we often desire to predict a new observation $y_s$ at $s$.  For instance, we might use a generalized linear model \citep{McCullaghNelder1989} or one of its extensions to predict species density or occurrence in a new location, or to predict the future trend of a population.

Early in their training, ecologists and statisticians are warned against extrapolating statistical relationships
past the range of observed data.  This caution is easily interpreted in the context of single-variable linear regression analysis; one should be cautious in using the fitted relationship to make predictions at some new point $y_s$ whenever $x_s < \min({\bf x})$ or $x_s > \max({\bf x})$.  But what about more complicated situations
where there are multiple explanatory variables, or when one uses a spatial regression model to account for the residual spatial autocorrelation that is inevitably present in patchy ecological data \citep{LichsteinEtAl2002}?  How reliable are spatially- or temporally-explicit predictions in sophisticated models for animal abundance and occurrence?

@@

Our aim in this paper is to investigate one possibility for defining extrapolation in the generalized linear model and its extensions, including generalized additive models \citep[GAMs;][]{HastieTibshirani1999,Wood2006} and spatial, temporal, or spatio-temporal regression models (STRMs).  In particular, we exploit some of the same ideas used in multiple linear regression regarding leverage and outliers \citep{Cook1979} to operationally define ``extrapolation" as making predictions that occur outside of a generalized independent variable hull (gIVH) of observed data points. Application of the gIVH and related criterion (e.g. prediction variance) can provide intuition regarding the reliability of predictions in unobserved locations, and can aid in model construction and survey design. We illustrate use of the gIVH on simulated count data, on a species distribution model (SDM) for ribbon seals ({\it Histriophoca fasciata}) in the eastern Bering Sea, and on a population trend model for Steller Sea Lions ({\it Phoca fasciata}).

@@

We simulated two sets of design data to help illustrate application of the IVH (Fig. \ref{fig:IVH}).  In simple linear regression with one predictor variable, predictions on a hypothetical response variable obtained at covariate values below the lowest observed value or above the highest observed value are primarily outside of the IVH.  We suspect this result conforms to most ecologists intuition about what constitutes ``extrapolating past the observed data."  However, fitting a quadratic model exhibits more nuance; if there is a large gap between design points, it is entirely possible that intermediate covariate values will also be outside of the IVH and thus more likely to result in problematic predictions.  Fitting a model with two covariates and both linear and quadratic effects, the shape of the IVH is somewhat more irregular, and even includes a hole in the middle of the surface when interactions are modeled (Fig. \ref{fig:IVH}).  These simple examples highlight the sometimes counterintuitive nature of predictive inference, a problem that can only become worse as models with more dimensions are contemplated (including those with temporal or spatial structure).  Fortunately, the ideas behind the IVH provide a potential way forward.

Cook's (1979) formulation for the IVH is particular to linear regression analysis, which assumes iid Gaussian error. Thus, it is not directly applicable to generalized models, such as those including alternative error distributions (e.g., Poisson, binomial) or spatial random effects.  Further, the hat matrix is not necessarily well defined for models with more general spatial structure. However, since the hat matrix is proportional to prediction variance, Cook (1979) notes that design points with maximum prediction variance will be located on the boundary of the IVH.  We therefore define a generalized independent variable hull (gIVH) as the set of all potential design points $\mathcal{S}_0$ (note that $\mathcal{S}_0$ can include both observed and unobserved design points) such that
\begin{linenomath*}
\begin{equation}
  \label{eq:gIVH}
  {\rm var}(\lambda_s) \le {\rm max} [{\rm var}(\boldsymbol{\lambda}_{\mathcal{S}})],
\end{equation}
\end{linenomath*}
where $s \in \mathcal{S}_0$, $\lambda_s$ corresponds to the mean prediction at $s$, $\mathcal{S}$ denotes the set of observed design points, and $\boldsymbol{\lambda}_\mathcal{S}$ denotes predictions at $\mathcal{S}$.

Generalizations of the linear model are often written in the form
\begin{linenomath*}
\begin{equation}
  Y_s \sim f_Y(g(\mu_s)),
\end{equation}
\end{linenomath*}
where $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives an inverse link function, and
$\mu_s$ is a linear predictor.  For many such generalizations, it is possible to specify the $\mu_s$ as
\begin{linenomath*}
\begin{equation}
  \boldsymbol{\mu} = {\bf X}_{aug}\boldsymbol{\beta}_{aug} + \boldsymbol{\epsilon},
  \label{eq:mu-aug}
\end{equation}
\end{linenomath*}
where the $\boldsymbol{\epsilon}$ represent Gaussian error, ${\bf X}_{aug}$ denotes an augmented design matrix, and $\boldsymbol{\beta}_{aug}$ denote an augmented vector of parameters.  For instance, in a spatial model, $\boldsymbol{\beta}_{aug}$ might include both fixed effect parameters and spatial random effects in a reduced dimension subspace (see Appendix A for examples of how numerous types of models can be written in this form).

When models are specified as in Eq. \ref{eq:mu-aug}, we can write prediction variance generically as
\begin{linenomath*}
\begin{equation}
  \label{eq:var-mu}
  {\rm var}(\boldsymbol{\mu}_x)= {\bf x} {\rm var}(\hat{\boldsymbol{\beta}}_{aug}) {\bf x}^\prime,
\end{equation}
\end{linenomath*}
where it is understood that the exact form of ${\bf x}$ and ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the model chosen (i.e., GLM, GAM, or STRM; Appendix A). This expression for prediction variance is on the linear predictor scale; if a non-identity link function is used, we can use the delta method \citep{Dorfman1938,VerHoef2012} to approximate prediction variance on the real scale (i.e. the scale in which data are measured) as
\begin{linenomath*}
\begin{eqnarray}
  \label{eq:var-lambda}
  {\rm var}(\boldsymbol{\lambda}_x) & = & {\rm var}(g(\boldsymbol{\mu}_x)) \\
  & \approx & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x)\boldsymbol{\Delta}^\prime,
%  \boldsymbol{\Delta} & = & \left[ \begin{array}{cccc}
%      \frac{\partial g(\mu_1)}{\partial \mu_1} & \frac{\partial g(\mu_1)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_1)}{\partial \mu_n} \\
%      \frac{\partial g(\mu_2)}{\partial \mu_1} & \frac{\partial g(\mu_2)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_2)}{\partial \mu_n} \\
%      \vdots & \vdots & \ddots & \vdots \\
%      \frac{\partial g(\mu_n)}{\partial \mu_1} & \frac{\partial g(\mu_n)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_n)}{\partial \mu_n} \\
%  \end{array} \right].
\end{eqnarray}
\end{linenomath*}
where $\boldsymbol{\Delta}$ is a matrix of partial derivatives $\partial g(\mu_r)/\partial \mu_c$ ($r$ and $c$ denoting rows and columns of $\boldsymbol{\Delta}$, respectively).  Under common univariate link functions (e.g. log, logit, probit), $\boldsymbol{\Delta}$ has a diagonal form, while for multivariate links (e.g. multinomial logit) $\boldsymbol{\Delta}$ will be dense.

Unlike Cook's original formulation, the gIVH will often involve unknown parameters.  Further, the exact form of ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the underlying model structure and estimation procedure.  For most of the following treatment, we shall assume that data have already been collected (see Discussion for comments on the potential use of the gIVH in survey planning), and that informative probability distributions can be specified for these parameters.  Specifically, we impose a probability distribution $[\boldsymbol{\theta}]$ for unknown parameters, and compute prediction variance as
\begin{linenomath*}
\begin{equation}
  \label{eq:E_var_lambda}
  {\rm var}(\boldsymbol{\lambda}_x | {\bf Y}) = \int_{\boldsymbol{\theta}} {\rm var}(\boldsymbol{\lambda}_x | {\bf Y}, \boldsymbol{\theta}) [\boldsymbol{\theta}] d \boldsymbol{\theta}.
\end{equation}
\end{linenomath*}
Our approach in this paper is to conduct a Bayesian analysis and substitute the posterior distribution $[\boldsymbol{\theta}|{\bf Y}]$ for $[\boldsymbol{\theta}]$ in Eq. \ref{eq:E_var_lambda}.  However, one could also consider substituting parametric approximations based on output from standard likelihood-based analyses.

We propose to use the gIVH in much the same manner as Cook (1979).  In particular, we use the gIVH to differentiate whether spatial predictions are interpolations (predictive design points lying inside the gIVH) or extrapolations (predictive design points lying outside the gIVH).  The gIVH, together with posterior prediction variance, seem ideally situated to diagnosing potential extrapolation issues. Further, one can compare prediction variance in places one has data to places where predictions are desired to gauge the relative amount of information that predictions will be based on.

\subsection{Computing}

We developed a package \texttt{SpatPred} in the R statistical programming environment \citep{RTeam2012} to simulate data and conduct all analyses.  The seal and sea lion datasets are included as part of this package, and are available at \url{https://github.com/pconn/SpatPred}.  The R package has also been archived via figshare [WE PLAN TO DO THIS WHEN THE MANUSCRIPT IS ACCEPTED FOR PUBLICATION].

@@

We used simulation to investigate the usefulness of the gIVH for limiting the scope of inference when fitting models to count data that have already been collected.  We also applied the gIVH to examining the robustness of a species distribution model (SDM) of ribbon seals in the eastern Bering Sea, and to time series projections of Steller sea lion population trends.

@@

We conducted a simulation study to investigate whether the gIVH (and accompanying prediction variance) was useful in diagnosing prediction biases when analyzing animal count data. For each of 100 simulations, we generated animal abundance over a $30 \times 30$ grid assuming that animal density was homogeneous in each grid cell.  Animal abundance was generated as a function of three hypothetical spatially autocorrelated habitat covariates (Appendix B).   For each simulated landscape, we conducted virtual surveys of $n=45$ survey units using two different designs: (1) a spatially balanced sample using a random tesselation design \citep{StevensOlsen2004}, and (2) a convenience sample where the probability of sampling was greater for cells closer to a ``base of operations" located in the middle of the survey grid. The former approach preserves randomness while seeking a degree or regularity when distributing sampling locations across the landscape, while the latter may be easier to implement.

We configured virtual sampling quadrats such that they encompassed 10\% of the area of each selected grid cell.  For ease of presentation and analysis, we assumed detection probability was 100\% in each quadrat.  Once animal counts were simulated, three different estimation models were fitted to the data: a GLM, a GAM, and an STRM (Appendix A).  The GLM and STRM expressed log-density as a function of linear and quadratic covariate effects (as well as first order interactions), while the GAM used a kernel smoother with 4 knots for each covariate (Appendix A).  Each model was provided with two of the three covariates used to generate the data.

For each simulated data set and model structure, we calculated the posterior predictive variance and resulting $\text{gIVH}$ as in Eq. \ref{eq:gIVH}. We then calculated posterior predictions of animal abundance within and outside of each gIVH in order to gauge bias as a function of this restriction.  Specifically, the performance of $\text{gIVH}$ may help decide its utility in limiting the scope of inference once data have been collected and analyzed, and perhaps point out areas worthy of additional sampling.  A fuller, technical description of the simulation study design is provided in Appendix B; for illustration of a single simulation replicate see Fig. \ref{fig:sim_maps}.

Posterior predictions from simulations indicated that the distribution for proportional error in total abundance was right skewed when statistical inference was made with regard to the entire survey area (Fig. \ref{fig:sim_boxplots}).  This was particularly true for GLM and STRM models, and was exacerbated when convenience sampling was employed.  The magnitude of mean absolute bias was reduced when inference was constrained to the gIVH for all 6 configurations of survey design and estimation model.  Interestingly, the convolution kernel GAMs we employed often underestimated total abundance, perhaps because of dampening ``edge effects" at extreme ranges of observed covariates, or possibly because interactions among covariates were not modeled.  For GLMs and STRMs, positive proportional bias was the rule, and was of concerning magnitude (e.g. $\approx 0.3$; Fig. \ref{fig:sim_boxplots}) for GLMs and STRMs when convenience sampling was employed and inference was not restricted to the gIVH.

\subsection{Ribbon seal SDM}
As part of an international effort, researchers with the U.S. National Marine Fisheries Service conducted aerial surveys over the eastern Bering Sea in 2012 and 2013.  Agency scientists used infrared video to detect seals that were on ice, and collected simultaneous digital photographs to provide information on species identity. Here, we use spatially referenced count data from photographed ribbon seals, {\it Phoca fasciata} (Fig. \ref{fig:ribbon}) on a subset of 10 flights flown over the Bering Sea from April 20-27, 2012.  We limited flights to a one week period because sea ice melts rapidly in the Bering Sea in the spring, and modeling counts over a longer duration would likely require addressing how sea ice and seal abundance changes over both time and space \citep[see][]{ConnEtAl2015}. However, limiting analysis to a one week period makes the assumption of static sea ice and seal densities tenable \citep{ConnEtAl2014}.

Our objective with this dataset will be to model seal counts on transects through 25km by 25km grid cells as a function of habitat covariates and possible spatial autocorrelation. Estimates of apparent abundance can then be obtained by summing predictions across grid cells. Figure \ref{fig:flights} shows the transects flown and the number of ribbon seals encountered in each cell, and Figure \ref{fig:covs} show explanatory covariates gathered to help predict ribbon seal abundance.  These data are described in fuller detail by \citet{ConnEtAl2014}, who extend the modeling framework of STRMs to account for incomplete detection and species misidentification errors.  Since our focus in this paper is on illustrating spatial modeling concepts, we devote our efforts to the comparably easier problem of estimating apparent abundance (i.e., uncorrected for vagaries of the detection process).

Inspection of ribbon seal data (Fig. \ref{fig:flights}) immediately reveals a potential
issue with spatial prediction: abundance of ribbon seals appears to be maximized in the southern and/or southeast quadrant of the surveyed area.  Predicting abundance in areas further south and east may thus prove problematic, as the values of several explanatory covariates (Fig. \ref{fig:covs}) are also maximized in these regions.  To illustrate, let $Y_i$ denote the ribbon seal count ($Y_i$) obtained in sampled grid cell $i$.  Suppose that counts arise according to a log-Gaussian Cox process, such that
\begin{eqnarray}
  \label{eqn:Cox}
  Y_i \sim {\rm Poisson}(\lambda_i) \mbox{ and } \\
  \log(\lambda_i) = \log(P_i) + \log(A_i) + \theta_i + \epsilon_i, \nonumber
\end{eqnarray}
where $P_i$ gives the proportion of area photographed in grid cell $i$, $A_i$ gives the proportion of suitable habitat in cell $i$ (ribbon seals do not often utilize land), $\theta_i$ is a linear predictor, and $\epsilon_i$ is normally distributed $iid$ error.  By formulating $\theta_i$ differently, we can arrive at representations characteristic of GLMs, GAMs, and STRMs (see Appendix S1).

We start by fitting a hierarchical GLM, GAM, and STRM to the full suite of predictor covariates (Fig. \ref{fig:covs}).  For the GLM and STRM, we fit a model with linear effects of all predictor variables, and with an additional quadratic term for ice concentration \citep[seal density is often maximized at an intermediate value of ice concentration; see][]{VerHoef2013,ConnEtAl2014}. For the STRM, we imposed a restricted spatial regression (RSR) formulation for spatially autocorrelated random effects, where dimension reduction was accomplished by only selecting eigenvectors of the spectral decomposition associated with eigenvalues that were greater than 0.5 (see Appendix S1 for additional information on model structure).  For the GAM, we employed smooth terms with 8 knots for each covariate.  This included 4 knots at the minimum, 1/3 and 2/3 quantiles, and the maximum observed covariate values, in addition to 4 knots above and below the range of observed data to reduce the magnitude of biased endpoint behavior.

As with simulated data examples, we adopted a Bayesian perspective and conducted estimation using MCMC (see Appendix S1 for algorithm details) with 60,000 iterations where the first 10,000 iterations were discarded as a burn-in.  We generated posterior predictions of ribbon seal abundance across the landscape as
\begin{equation}
  N_i \sim {\rm Poisson}(A_i \lambda_i),
\end{equation}
and calculated the gIVH as in Eqn. \ref{eq:gIVH}, with delta method modifications as specified in Eqns. \ref{eq:var-mu}-\ref{eq:E_var_lambda}.
Initial spatial predictions using the three models (GLM, GAM, STRM) all produced extremely high, unbelievable predictions along the southern boundary of the study area \ref{fig:ribbon_plot}.  Predictions in this region were also largely out of the gIVH, indicating the potential utility for the gIVH in revealing problematic extrapolations.

We considered several possible alternatives for trying to obtain more robust abundance estimates before settling on a preferred alternative.  First, one could refine the study area to eliminate predictions outside of the gIVH (as in the simulation study).  However, this is not ideal in that one does not get an abundance estimate for the whole study area, and it may be difficult to compare abundance from one year to the next using this approach. Second, one could consider a simpler estimation model, such as one post-stratifying by ecoregion as was used in a previous analysis of ribbon seal data to try to avoid this problem \citep{ConnEtAl2014}.  Finally, one could build in a priori knowledge of habitat preferences into the model structure.  We adopted the latter solution, incorporating ``pseudo-absences" (i.e., zero counts where sampling was not conducted) in locations where it would have been (nearly) impossible to detect seals.  Specifically, we inserted pseudo-absences in cells where ice concentrations were $<0.1\%$.  This solution seemed the most logical, as many of the large, anomalous predictions were over open water along the southern edge of the study area, where we would have obtained zero counts had they been surveyed.
